Inferring neural networks from fluorescent calcium imaging using partial correlation and sample weighting \textcolor{red}{plus court}

### 2014-06-06 article.tex
add dot

 % change the arguments, as appropriate, in the following:
\jmlrvolume{1}
\jmlryear{2014}
\jmlrworkshop{Neural Connectomics Workshop}

\title[Connectomics challenge]{Inferring neural networks from fluorescent
                               calcium imaging using partial correlation and
                               sample weighting}

@@

@@

@@

Each neuron signal $i$ is a random continuous variable $X_i$,
which follows a probability density function $p_\mathcal{X_i}(x_i)$.
Two neurons $X_i$, $X_j$ will be said conditionally independent
to a set of neurons $X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$, denoted
by  $X_i \perp X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$,  if
\begin{align*}
&p_{X_i, X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_i, x_j | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\}) \\
=&p_{X_i | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_i | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\})
p_{X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
    (x_j | x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\})
\end{align*}
for all real numbers $x_1,\ldots,x_p $ such that
$p_{X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
(x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\}) > 0$. Note that this
reciprocal is true only under additional hypotheses.

Under the hypothesis that the neurons follow a jointly Gaussian distribution
$p_{X_1,\ldots,X_p}(x_1,\ldots,x_p) \sim \mathcal{N}(\mu; \Sigma)$
of mean vector $\mu$ and of covariance matrix $\Sigma$, two neurons $X_i$
and $X_j$ are independent $X_i \perp X_j$ if they have zero Pearson correlation
coefficient
\[
\rho_{X_iX_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii} \Sigma_{jj}}}
\]
and are conditionally independent to all others neurons
$X_i \perp X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$ if they
have zero partial correlation coefficient
\[
\rho_{X_i, X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}} =
\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}}.
\]

@@

\subsection{Causal discovery methodology}
% Introduce a custom solution to make the Y matrix asymmetric
% Introduce directivity
Since the partial correlation is a symmetric measure, the causal mechanism behind the
interaction ($A$ directly causes $B$, $B$ directly causes $A$ or both) can not
be inferred\footnote{Also note that two other causal mechanisms might be
implied while having a non-zero partial coefficient: (i) a pair of variables
is induced by a common (hidden) variable; (ii) $A$ (respectively $B$) is
conditionally correlated to a hidden variable affecting $B$ (resp. $A$)
\cite{de2004discovery}.} directly from a symmetric adjacency matrix.

By ignoring the directivity, we systematically
make a true positive and a false positive for direct link (see a more detailed
discussion about the chosen metric in section \ref{sec:metric}). Therefore,
performance might be improved by predicting an asymmetric adjacency matrix.
Hoping to orientate correctly the most obvious directed links, we attempt to
retrieve some causal information, i.e. the directivity of the links, by computing
and stacking with our symmetric adjacency matrix a matrix of pairwise activation
heuristics. This method computes an anti-symmetric activation score $z_{ij}$
based the variation of fluorescence signal of a neuron $j$ due to a neuron $i$
over all timesteps:
\[
z_{ij} = \sum_{t=1}^n I(X_{t+1,j} - X_{t, i} \in [f_1, f_2]) -  I(X_{t+1,i} - X_{t, j} \in [f_1, f_2])
\]
where $f_1$ and $f_2$ are parameters of the method and $I$ is the
indicator function. % TODO say that we use only the 3 first filters for this

@@

\section{Experiments}
% Datasets
To assess our method, we have used four datasets provided by the organizer
of the Connectomics challenge : normal-1, normal-2, normal-3, normal-4. Those
datasets are time series of size $n=179500$ of image calcium from simulated
neuronal network \cite{stetter2012model} with $p=1000$ neurons. The networks
to inferred have around 15000 arcs or a density of $0.15\%$.

@@

% Possible extensions
% partial autocorrelation function (sometimes "partial correlation function")
% conditional independence test
% Speak about the complexity and parallelization of the methods
\begin{scriptsize}
\paragraph{Acknowledgements.} A. Joly and G. Louppes are research
fellows of the FNRS, Belgium.  A. Sutera is supported by a F.R.I.A. scholarship.
This work is supported by PASCAL2 and the IUAP
DYSCO, initiated by the Belgian State, Science Policy Office.
Computational ressources have been provided by the Consortium des Équipements
de Calcul Intensif (CÉCI), funded by the Fonds de la Recherche Scientifique de
Belgique (F.R.S.-FNRS) under Grant No. 2.5020.11
\end{scriptsize}

>>>>

 % change the arguments, as appropriate, in the following:
\jmlrvolume{1}
\jmlryear{2014}
\jmlrworkshop{\textcolor{red}{Neural Connectomics Workshop}}

\title[Connectomics challenge]{Inferring neural networks from fluorescent
                               calcium imaging using partial correlation and
                               sample weighting \textcolor{red}{plus court}}

Simple and robust inference of connectomes using partial correlation coefficients

%% 
Discovery/Infering/Retrieving/... connectomes 
Robust
Using partial correlation
[from calcium imaging]

\textcolor{red}{Discussion on the order of the partial correlation}

@@

%% Proposition (sujet à changements : on en discutera):
% Antonio
% {Aaron, Arnaud, Vincent} : aucune préférence pour ma part
% Gilles
% Damien?
% Pierre?

@@

\textcolor{red}{Motiviation pour l'inverse : retirer les liens indirects} \\

@@

Let us consider that each neuron signal $i$ is a random continuous variable $X_i$
which follows a probability distribution $p_{X_i}$ and let us denote by $p_{X_i,X_j|\mathcal{Z}}$ the 
joint probability distribution of neurons $X_i$ and $X_j$ conditioned on a subset of variables $\mathcal{Z}$ (eventually empty). Let us define $\mathcal{V}$
the set of all neurons $X_1, ..., X_p$ and $\mathcal{V}^{-i,j}$ the same set without
neurons $X_i$ and $X_j$.

Each neuron signal $i$ is a random continuous variable $X_i$,
which follows a probability density function $p_\mathcal{X_i}(x_i)$.
Two neurons $X_i$, $X_j$ will be said conditionally independent
to a set of neurons $\mathcal{V}^{-i,j}$, denoted
by  $X_i \perp X_j | \mathcal{V}^{-i,j}$,  if $p_{X_i,X_j|\mathcal{V}^{-i,j}} = p_{X_i|\mathcal{V}^{-i,j}}  p_{X_j|\mathcal{V}^{-i,j}} $.
%for all real numbers $x_1,\ldots,x_p $ such that
%$p_{X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}}
%(x_1,\ldots,x_p \backslash \left\{x_i, x_j\right\}) > 0$. Note that this
%reciprocal is true only under additional hypotheses.

Under the hypothesis that the neurons follow a jointly Gaussian distribution
$p_{X_1,\ldots,X_p}(x_1,\ldots,x_p) \sim \mathcal{N}(\mu; \Sigma)$
of mean vector $\mu$ and of covariance matrix $\Sigma$, two neurons $X_i$
and $X_j$ are independent $X_i \perp X_j$ if the Pearson correlation
coefficient
\[
\rho_{X_iX_j} = \frac{\Sigma_{ij}}{\sqrt{\Sigma_{ii} \Sigma_{jj}}} = 0,
\] 
and are conditionally independent to all others neurons
$X_i \perp X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}$ if the partial correlation coefficient
\[
\rho_{X_i, X_j | X_1,\ldots,X_p \backslash \left\{X_i, X_j\right\}} =
\frac{\Sigma^{-1}_{ij}}{\sqrt{\Sigma^{-1}_{ii} \Sigma^{-1}_{jj}}} = 0.
\]

@@

\subsection{Causal discovery methodology}
% Introduce a custom solution to make the Y matrix asymmetric
% Introduce directivity
Since the partial correlation is a symmetric measure, the causal mechanism behind the
interaction ($A$ directly causes $B$, $B$ directly causes $A$ or both) can not
be directly inferred\footnote{Also note that two other causal mechanisms might be
implied while having a non-zero partial coefficient: (i) a pair of variables
is induced by a common (hidden) variable; (ii) $A$ (respectively $B$) is
conditionally correlated to a hidden variable affecting $B$ (resp. $A$)
\cite{de2004discovery}.} from a symmetric adjacency matrix.

By ignoring the directivity, we systematically
make a true positive and a false positive for direct link (see a more detailed
discussion about the chosen metric in section \ref{sec:metric}). Therefore,
performance might be improved by predicting an asymmetric adjacency matrix.
Hoping to orientate correctly the most obvious directed links, we attempt to
retrieve some causal information, i.e. the directivity of the links, by computing
and stacking with our symmetric adjacency matrix a matrix of pairwise activation
heuristics. This method computes an anti-symmetric activation score $z_{ij}$
based the variation of fluorescence signal of a neuron $j$ due to a neuron $i$
over all timesteps:
\[
z_{ij} = \sum_{t=1}^n 1(X_{t+1,j} \ge X_{t, i} \in [f_1, f_2]) -  1(X_{t+1,i} \ge X_{t, j} \in [f_1, f_2])
\] \textcolor{red}{to rewrite}
where $f_1$ and $f_2$ are parameters of the method and $I$ is the
indicator function. % TODO say that we use only the 3 first filters for this

@@

\section{Experiments}

% Datasets
To assess our method, we have used four datasets provided by the organizer
of the Connectomics challenge : normal-1, normal-2, normal-3, normal-4. Those
datasets are time series of size $n=179500$ of image calcium from simulated
neuronal network \cite{stetter2012model} with $p=1000$ neurons. The networks
to inferred have around 15000 arcs or a density of $0.15\%$ and are similar to those
used to rank challenge participants.

Note that our parameter tuning was mainly on \textit{normal-1} dataset and
partly on \textit{normal-4} dataset, therefore it may justify why these
networks are better retrieved despite the fact that some might be simpler to
infer.

@@

% Possible extensions
% partial autocorrelation function (sometimes "partial correlation function")
% conditional independence test
% Speak about the complexity and parallelization of the methods

\begin{scriptsize}

\paragraph{Acknowledgements.} A. Joly and G. Louppe are research fellows of
the FNRS, Belgium. We also thank Damien for pizzas.  A. Sutera is supported by
a F.R.I.A. scholarship.

\end{scriptsize}
